Manuscript Figures for Genome Biology

### 2018-06-05 datasets/RNA-Seq/yeast-simulation.Rmd

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param qq logical whether to return qqplot of test statistics (for diagnosing
#'  problems with Scott method)
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd, uninformativeCovariate = FALSE, qq = FALSE){

@@

    # randomly sample a log2FC from original FCs (without regard to DE)
    log2FC <- rep(0, nrow(dds_test))
    log2FC[DE] <- resFULL$log2FoldChange[DE]

    counts_new <- counts(dds_test)
    counts_new[,colData(dds_test)$fake=="A"] <- counts(dds_test)[,colData(dds_test)$fake=="A"] * 2^log2FC
    counts_new <- apply(counts_new, 2, as.integer)
    counts(dds_test) <- counts_new
  }

@@

  if(qq){
    return(ggplot(data.frame(resTEST), aes(sample = stat)) + 
             geom_qq() +
             geom_abline())
  }

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE, 
                               ind_covariate = ifelse(uninformativeCovariate, 
                                                      rnorm(nrow(resTEST)),
                                                      resTEST$baseMean),
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test))  ))

  # filter NAs and those with less than 50% expressed 
  geneExp <- geneExp %>% na.omit() %>% dplyr::filter(pzero < 0.5)

@@

  assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
  pf <- estimatePerformanceMetrics(sb, tidy=TRUE, alpha=seq(0.01, 0.1, 0.01))
pf %>% filter(performanceMetric %in% c("rejections", "FDR", "TPR")) %>%
  with(table(is.na(value), blabel, performanceMetric))

  rowData(sb)$log2FC <- geneExp$effect_size

@@

if (!file.exists(resfile_n5)){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null5, file = resfile_n5)
}else{
  null5 <- readRDS(file = resfile_n5)
}

# qqplots
plotfile <- file.path(datdir, "null5_qqplots.pdf")
if (!file.exists(plotfile)){
  qqs <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   qq =TRUE)
  failed <- ifelse(sapply(null5, 
                          function(x) sum(is.na(assays(x)$qvalue[,22])) > 0),
                   "Failed", "OK")
  pdf(plotfile, width=4, height=4)
  for(i in 1:length(qqs)){
    print(qqs[[i]] +
      ggtitle(paste0(failed[i])))
  }
  dev.off()
}
```

@@

# qqplots
plotfile <- file.path(datdir, "null10_qqplots.pdf")
if (!file.exists(plotfile)){
  qqs <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   qq =TRUE)
  failed <- ifelse(sapply(null10, 
                          function(x) sum(is.na(assays(x)$qvalue[,22])) > 0),
                   "Failed", "OK")
  pdf(plotfile, width=4, height=4)
  for(i in 1:length(qqs)){
    print(qqs[[i]] +
      ggtitle(paste0(failed[i])))
  }
  dev.off()
}
```

@@

Here we'll repeat the above, but for a DE comparison 5 versus 5 Snf2 samples,
where the groups are selected randomly and 500 DE genes are added. 
This will be done for 100 random splits.

@@

```{r d5, results='hide', message=FALSE}
rseed <- 198
sampleSize <- 5
nDE <- 500

@@

# qqplots
plotfile <- file.path(datdir, "de5_qqplots.pdf")
if (!file.exists(plotfile)){
  qqs <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   qq =TRUE)
  failed <- ifelse(sapply(de5, 
                          function(x) sum(is.na(assays(x)$qvalue[,22])) > 0),
                   "Failed", "OK")
  pdf(plotfile, width=4, height=4)
  for(i in 1:length(qqs)){
    print(qqs[[i]] +
      ggtitle(paste0(failed[i])))
  }
  dev.off()
}
```

@@

Here we'll repeat the above, but for a DE comparison 10 versus 10 Snf2 samples,
where the groups are selected randomly and 500 DE genes are added.
This will be done for 100 random splits.

@@

```{r d10, results='hide', message=FALSE}
rseed <- 961
sampleSize <- 10
nDE <- 500

@@

# qqplots
plotfile <- file.path(datdir, "de10_qqplots.pdf")
if (!file.exists(plotfile)){
  qqs <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   qq =TRUE)
  failed <- ifelse(sapply(de10, 
                          function(x) sum(is.na(assays(x)$qvalue[,22])) > 0),
                   "Failed", "OK")
  pdf(plotfile, width=4, height=4)
  for(i in 1:length(qqs)){
    print(qqs[[i]] +
      ggtitle(paste0(failed[i])))
  }
  dev.off()
}
```

@@

```{r d5-uc, results='hide', message=FALSE}
rseed <- 198
sampleSize <- 5
nDE <- 500

@@

```{r d10-uc, results='hide', message=FALSE}
rseed <- 961
sampleSize <- 10
nDE <- 500

>>>>

```{r}
#' @param X simnumber
#' @param seed random seed
#' @param dds is the full DESeq2 dataset
#' @param resFULL is the results object from running DESeq2 on dds (used in
#'   to obtain the fold change effect sizes when adding non-null genes)
#' @param sampleSize is the number of samples in each condition
#' @param nDE is the number of DE genes
#' @param bd is the bench design object
#' @param BPPARAM is the BiocParallel bpparam argument to pass to DESeq. Since
#'  parallelization happens by forking multiple instances of this function, we 
#'  will specify only one worker per DEseq intance.
#' @param uninformativeCovariate logical indicating whether to replace overall mean 
#'  independent covariate with a randomly generated (uninformative) covariate. 
#'  Default is FALSE.
#' @param pvalHists logical whether to return histograms of pvalues by covariate
#'  instead of SB objects
simulateOneSplit <- function(X, rseed, dds, resFULL, nDE, sampleSize,
                             bd, uninformativeCovariate = FALSE, pvalHists = FALSE){

@@

    # randomly sample a log2FC from original FCs (without regard to DE)
    counts_new <- counts(dds_test)
    log2FC <- rep(0, nrow(dds_test))
    log2FC[DE] <- resFULL$log2FoldChange[DE]

    # randomize which condition is shifted up or down
    ran <- runif(nrow(dds_test)) 
    multiplier <- ifelse(ran < 0.5, 2^log2FC, 1/(2^log2FC))
    refcond <- ifelse(ran < 0.5, "A", "B")
    down <- which(ran < 0.5)

    counts_new[down,colData(dds_test)$fake==unique(refcond[down])] <- 
        counts(dds_test)[down, colData(dds_test)$fake==unique(refcond[down])] *
        multiplier[down]
    counts_new[-down,colData(dds_test)$fake==unique(refcond[-down])] <- 
        counts(dds_test)[-down, colData(dds_test)$fake==unique(refcond[-down])] *
        multiplier[-down]

    counts_new <- apply(counts_new, 2, as.integer)

    counts(dds_test) <- counts_new
  }

  # replace existing size factors 
  dds_test <- estimateSizeFactors(dds_test)

@@

  geneExp <- tbl_df(data.frame(geneName=rownames(resTEST), 
                               pval=resTEST$pvalue, 
                               SE=resTEST$lfcSE,                 
                               ind_covariate = resTEST$baseMean,
                               effect_size = resTEST$log2FoldChange, 
                               test_statistic = resTEST$stat,
                               qvalue = truth,
                               pzero = rowSums(counts(dds_test)==0)/
                                       ncol(counts(dds_test)))) %>%
                dplyr::filter(!is.na(pval))

  if(uninformativeCovariate)
    geneExp$ind_covariate <- rnorm(nrow(geneExp))

   if(pvalHists){
    return(strat_hist(geneExp, pvalue="pval", covariate="ind_covariate", maxy =10))
  }

@@

  assayNames(sb) <- "qvalue"
  sb <- addDefaultMetrics(sb)
  rowData(sb)$log2FC <- geneExp$effect_size

@@

if (!file.exists(resfile_n5)){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd)
  saveRDS(null5, file = resfile_n5)
}else{
  null5 <- readRDS(file = resfile_n5)
}
```

@@

@@

Here we'll repeat the above, but for a DE comparison 5 versus 5 Snf2 samples,
where the groups are selected randomly and 2000 DE genes are added. 
This will be done for 100 random splits.

@@

```{r d5, results='hide', message=FALSE}
rseed <- 198
sampleSize <- 5
nDE <- 2000

@@

# pvalue histograms
plotfile <- file.path(datdir, "de5_pvalhists.pdf")
if (!file.exists(plotfile)){
  hists <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   pvalHists = TRUE)
  pdf(plotfile, width=8, height=4)
  for(i in 1:length(hists)){
    print(hists[[i]])
  }
  dev.off()
}
```

@@

Here we'll repeat the above, but for a DE comparison 10 versus 10 Snf2 samples,
where the groups are selected randomly and 2000 DE genes are added.
This will be done for 100 random splits.

@@

```{r d10, results='hide', message=FALSE}
rseed <- 961
sampleSize <- 10
nDE <- 2000

@@

# pvalue histograms
plotfile <- file.path(datdir, "de10_pvalhists.pdf")
if (!file.exists(plotfile)){
  hists <- bplapply(X=1:B, FUN=simulateOneSplit, rseed=rseed, dds=dds, 
                   resFULL=resFULL, nDE=nDE, sampleSize=sampleSize, bd=bd, 
                   pvalHists = TRUE)
  pdf(plotfile, width=8, height=4)
  for(i in 1:length(hists)){
    print(hists[[i]])
  }
  dev.off()
}
```

@@

```{r d5-uc, results='hide', message=FALSE}
rseed <- 198
sampleSize <- 5
nDE <- 2000

@@

```{r d10-uc, results='hide', message=FALSE}
rseed <- 961
sampleSize <- 10
nDE <- 2000

### 2018-06-05 manuscript_figures/manuscript_figures.Rmd

```{r Figure2}
objects <- list.files( path, recursive=TRUE, pattern="rds", full.names=TRUE )
resfile <- objects[grepl("yeast", objects) & 
                   grepl("de5", objects) & 
                  !grepl("uninfCov", objects)] 
de5 <- readRDS(file=resfile)
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

@@

# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|null|polyester"), dataset) &
            !grepl("scRNAseq", casestudy) &
            !grepl("_uninfCov", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|null|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

# null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|de|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|yeast|null|_uninfCov"), dataset) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

>>>>

```{r Figure2}
objects <- list.files( path, recursive=TRUE, pattern="rds", full.names=TRUE )
resfile <- objects[grepl("yeast", objects) & 
                   grepl("de5", objects) & 
                  !grepl("uninfCov", objects)] 
                  !grepl("uninfCov", objects)] 
de5 <- readRDS(file=resfile)
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
res5d <- plotsim_standardize(de5, alpha = seq(0.01, 0.10, 0.01))

@@

# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|null|polyester"), dataset) &
            !grepl("scRNAseq", casestudy) &
            !grepl("uninfCov", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
# non-null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|null|uninfCov"), objects) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

# null
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|polyester|de|uninfCov"), objects) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))

@@

```{r}
wcs <- which(grepl("RNAseq", casestudy) & 
            !grepl(c("brain|mir|yeast|null|uninfCov"), objects) &
            !grepl("scRNAseq", casestudy))
objs <- objects[wcs]
labs <- gsub(".rds", "", 
             unlist(lapply(str_split(dataset[wcs], pattern = "-results-"), 
                      function(x) x[2])))
